---
title: "Wild boars around here"
output: html_notebook
---

```{r start, message=FALSE}
library(readxl)
library(dplyr)
library(naniar)
library(lubridate)
library(dygraphs)
library(leaflet)
library(xts)
```

Inspired by recent news on a [fortcoming book by Hannah Lutz](https://www.hbl.fi/artikel/vildsvin-som-vag-till-sjalvkannedom/), and the notebook on [Raccoon dog in Sweden](https://github.com/christopherkullenberg/AnalysisportalNotebooks/blob/master/mardhund.ipynb) by Christopher Kullenberg, I decided to have a look on how the wild boar (vildsvin in Swedish, villisika in Finnish) is doing around here. Based on a few stays in Italian and French countryside, I can confirm that yes, there are wild boars in mainland Europe. In Finland, I haven't seen a single one yet. Perhaps they wander along the Eastern border?

While at it, I decided to make use of the new R Notebook by RStudio. Having done a little R Markdown'ing before, watching [a helpful 45 min video tutorial](https://www.rstudio.com/resources/webinars/introducing-notebooks-with-r-markdown/) gave a head start.

First I went to [The Analysis portal for biodiversity data](https://www.analysisportal.se) like Christopher did, and downloaded data on this *Sus scrofa*.

```{r load, warning=FALSE}
data <- read_excel("SpeciesObservations.xlsx", sheet = "SLW Data")
nrow(data)
colnames(data)
```

The number of different variables in this citizen science data is impressive. Yet, seemingly many empty cells. What data should I expect NOT to find? To answer this question, the brand new [naniar package](https://github.com/njtierney/naniar) is of great help. 

Let's take the first half of the columns first, and sort them by missingness.

```{r missing1}
vis_miss(data[ , 1:as.integer(ncol(data)/2)], sort_miss = TRUE) + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 48, 
        vjust = 1, hjust = 1, size = 7))

```

And then the rest.

```{r missing2}
vis_miss(data[ , as.integer(ncol(data)/2+1):ncol(data)], sort_miss = TRUE) + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 48, 
        vjust = 1, hjust = 1, size = 7))
```

It looks like the dataset is a little patchy on extra information like the number of observed animals, what they were doing etc. This is understandable. On the other hand, the core details if you like such as location and date are well represented. I'd guess that some of these fields are mandatory.

Next we need to transform dates, and then we can count daily statistics.

```{r timeline}
data$Start <- as.Date(data$Start, format = "%Y-%m-%d")

dailysums <- data %>%
  rename(Date = Start) %>% 
  group_by(Date) %>% 
  summarise(Observations = n()) 
  
dailysums <- as.data.frame(dailysums, stringsAsFactors = FALSE)

# Pad dates
alltime <- seq.Date(min(dailysums$Date), max(dailysums$Date), by="day")
alltime_df <- data.frame(Date = alltime)
alltime_dailysums <- full_join(alltime_df, dailysums, by=c("Date"="Date"))
alltime_dailysums$Observations <- ifelse(is.na(alltime_dailysums$Observations), 0, alltime_dailysums$Observations)

rownames(alltime_dailysums) <- alltime_dailysums[[1]]
dailysums.xts <- as.xts(alltime_dailysums, dateFormat = "Date")

dygraph(dailysums.xts) %>%
  dyHighlight(highlightCircleSize = 5, 
              highlightSeriesBackgroundAlpha = 0.2,
              hideOnMouseOut = FALSE) %>% 
  # https://www.timeanddate.com/holidays/sweden/midsummer-day
  dyEvent("2010-06-26", "Midsummer", labelLoc = "top") %>% 
  dyEvent("2011-06-25", "Midsummer", labelLoc = "top") %>% 
  dyEvent("2012-06-23", "Midsummer", labelLoc = "top") %>% 
  dyEvent("2013-06-22", "Midsummer", labelLoc = "top") %>% 
  dyEvent("2014-06-21", "Midsummer", labelLoc = "top") %>% 
  dyEvent("2015-06-20", "Midsummer", labelLoc = "top") %>% 
  dyEvent("2016-06-25", "Midsummer", labelLoc = "top") %>%
  dyEvent("2010-12-24", "Christmas", labelLoc = "top") %>% 
  dyEvent("2011-12-24", "Christmas", labelLoc = "top") %>% 
  dyEvent("2012-12-24", "Christmas", labelLoc = "top") %>% 
  dyEvent("2013-12-24", "Christmas", labelLoc = "top") %>% 
  dyEvent("2014-12-24", "Christmas", labelLoc = "top") %>% 
  dyEvent("2015-12-24", "Christmas", labelLoc = "top") %>% 
  dyEvent("2016-12-24", "Christmas", labelLoc = "top") %>%
  dyOptions(stepPlot=FALSE,
            connectSeparatedPoints = FALSE,
            colors = RColorBrewer::brewer.pal(3, "Set2")) 


```

Select a region to zoom in. This way, you can contemplate on whether the sightings happened to occur around bigger public holidays, say.

Where in Sweden have people seen these animals? Was it a herd, or just few individuals? What were they doing? 

```{r map}

data <- data %>% 
  mutate(Lat = as.numeric(gsub(",", ".", DecimalLatitude))) %>% 
  mutate(Lon = as.numeric(gsub(",", ".", DecimalLongitude))) %>% 
  mutate(Time = EventTime) %>% 
  mutate(Count = ifelse(is.na(IndividualCount), 0, as.integer(IndividualCount))) %>%
  mutate(Quantity = ifelse(is.na(Quantity), "N/A", Quantity)) %>%
  mutate(LifeStage = ifelse(is.na(LifeStage), "N/A", LifeStage)) %>% 
  mutate(Behavior = ifelse(is.na(Behavior), "N/A", Behavior)) %>% 
  mutate(Sex = ifelse(is.na(Sex), "N/A", Sex)) %>% 
  mutate(OccurrenceRemarks = ifelse(is.na(OccurrenceRemarks), "N/A", OccurrenceRemarks))

max <- data %>% 
  arrange(desc(Count)) %>% 
  head()

leaflet(data) %>% 
  addTiles() %>%
  addCircles(lng = ~Lon, lat = ~Lat, weight = 2,
    radius = ~sqrt(Count) * 30, popup = paste(sep = "<br/>", 
                                          paste0("County: ",data$County), 
                                          paste0("Date: ", data$Start), 
                                          paste0("Time: ", data$Time),
                                          paste0("Quantity: ", data$Quantity),
                                          paste0("Life stage: ", data$LifeStage),
                                          paste0("Behavior: ", data$Behavior), 
                                          paste0("Sex: ", data$Sex), 
                                          paste0("Remarks: ", data$OccurrenceRemarks))) %>% 
  addMarkers(lng = max$Lon, lat = max$Lat, popup = paste(sep = "<br/>", 
                                          paste0("County: ",max$County), 
                                          paste0("Date: ", max$Start), 
                                          paste0("Time: ", data$Time),
                                          paste0("Quantity: ", max$Quantity),
                                          paste0("Life stage: ", max$LifeStage),
                                          paste0("Behavior: ", max$Behavior), 
                                          paste0("Sex: ", max$Sex), 
                                          paste0("Remarks: ", max$OccurrenceRemarks)))
```

Some of the observed groups are astonishingly big. To let these stand out on the map, I put a blue marker on those locations. One of them is in the outskirts of Stockholm, the capital of Sweden! From the popup we learn that here, the number of animals is due to free lunch. They are fed.

The most Northern observation comes near Umeå. 

Finland? This part is yet to be done. First I need to find some open data.
